Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model

### 2025-12-24 articles/arxiv/paper.tex
minor tweaks to allow arxiv submission

\graphicspath {{../../output/figures/}}

@@

      \begin{table}
        \footnotesize
        \center
        \input{../../output/tables/model_summary.tex}
        \caption{\it Summary of estimated runner outcome probability models. We estimated four different generalized (binomial) linear mixed-effects models for pickoff (PO) attempt and success and for stolen base (SB) attempt and success. For fixed effects we report the estimated effect with one standard error, and for random effects we report the estimated standard deviation term. For 2022, we encoded disengagements as always zero to reflect no pickoff limit.}
        \label{tab:model-summary}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_count_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-two-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_outs_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-two-player}
      \end{table}

      We observe that the lead distance generally (although not strictly) shortens as the count deepens (i.e., as the numbers of balls and strikes increase). This makes intuitive sense because the plate appearance is closer to ending, which relieves the pressure on the pitcher to avoid attempting a pickoff. The lead distance generally (although not strictly) increases as the number of outs increases. This makes intuitive sense because the runner is less likely to score without more aggressive action. In other words, the riskier lead distance is more palatable because the runner has a lower baseline probability of scoring from first base with two outs. The non-monotonicity of the optimal lead distance policy may be a consequence of estimating transition probabilities using empirical data.

@@

      \begin{table}[b]
        \centering
        \input{../../output/tables/lead_by_count_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_outs_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \input{../../output/tables/lead_by_players.tex}
        \caption{
          \it Optimal lead distance (in feet) by battery/runner skill and prior disengagements, for an 0-0 count with no outs and third base empty. For example, the hypothetical 90$^{th}$-percentile battery represents the 90$^{th}$ percentile at attempting pickoffs, successfully executing pickoffs, suppressing stolen base attempts, and catching would-be base stealers.
        }
        \label{tab:lead-by-players}
      \end{table}

@@

    The next result states and proves the policy iteration algorithm for our two-player zero-sum sequential game.
       \begin{proposition}
           \label{lem:policy_iteration}
           Consider the sequential game defined in Section \ref{sec:sg-model} and suppose Assumptions \ref{assum:game_halts} and \ref{assum:transition_pby_continuous} hold. Starting from any $\pi_\text{R}^0 \in \Pi_\text{R}$, define the following updates for all $k \geq 0$.
           \begin{align*}
              % \label{eqn:update-value-two-player}
              V^{\pi_{\text{R}}^k}(s) &\defeq \min_{\pi_\text{P} \in \Pi_\text{P}} V^{\pi_\text{R}^k,\pi_\text{P}}(s) && \forall s \in \mathcal{S}, \mbox{ and}\\
              \label{eqn:update-policy-two-player}
              \pi_\text{R}^{k+1}(s) &\in \argmax_{a_\text{R} \in A_\text{R}(s)} \left\{\min_{a_\text{p} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in \mathcal{S}}[r(s'|s) + V^{\pi_{\text{R}}^k}(s')] p(s'|s,a_\text{R},a_\text{P})\right\} && \forall s \in \mathcal{S}.
            \end{align*}
            Then $V^{\pi_\text{R}^k} \uparrow V^*$. 
       \end{proposition}
       \begin{proof}
           For any $\pi_\text{R} \in \Pi_\text{R}$, define the operator $T_{\pi_\text{R}}$ on $\mathbb{R}^{\abs{\cal S}}$ as 
           \begingroup
           \allowdisplaybreaks
           \begin{equation}
               \label{eqn:policy_iteration_temp1}
               (T_{\pi_\text{R}} u )(s) \defeq \min_{a_\text{P} \in A_\text{P}(s,\pi_\text{R}(s))}\sum_{s' \in S}\left[r(s'|s) + u(s')\right]p(s'|s,\pi_\text{R}(s),a_\text{P}) \ \forall s \in {\cal S}.
           \end{equation}
           Since $\pi_\text{R}$ is stationary, the best response of the pitcher corresponding to $\pi_\text{R}$ reduces to solving a MDP. Denote the value of this MDP as $V^{\pi_\text{R}}$, i.e., $V^{\pi_\text{R}}(s) \defeq \min_{\pi_\text{P} \in \Pi_\text{P}} V^{\pi_\text{R},\pi_\text{P}}(s) \ \forall s \in {\cal S}$. Using similar arguments as in the proof of Theorem  \ref{lem:vi_convergence}, it follows that $\lim_{k \to \infty}T_{\pi_\text{R}}^k u = V^{\pi_\text{R}}$, where $T_{\pi_\text{R}}^k$ is the composition of $T_{\pi_\text{R}}$ taken with itself $k$ times.  Also, using arguments similar to the proof of Proposition  \ref{lem:bellman_equations}, it follows that $V^{\pi_\text{R}}$ is the unique fixed point of $T_{\pi_\text{R}}$.

>>>>

\graphicspath {{figures/}}
\newcommand{\tablepath}{tables/}
\newcommand{\inputtable}[1]{\input{\tablepath #1}}

@@

      \begin{table}
        \footnotesize
        \center
        \inputtable{model_summary.tex}
        \caption{\it Summary of estimated runner outcome probability models. We estimated four different generalized (binomial) linear mixed-effects models for pickoff (PO) attempt and success and for stolen base (SB) attempt and success. For fixed effects we report the estimated effect with one standard error, and for random effects we report the estimated standard deviation term. For 2022, we encoded disengagements as always zero to reflect no pickoff limit.}
        \label{tab:model-summary}
      \end{table}

@@

      \begin{table}
        \centering
        \inputtable{lead_by_count_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-two-player}
      \end{table}

@@

      \begin{table}
        \centering
        \inputtable{lead_by_outs_two_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the two-player sequential game, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-two-player}
      \end{table}

      We observe that the lead distance generally, although not strictly, shortens as the count deepens (i.e., as the numbers of balls and strikes increase). This makes intuitive sense because the plate appearance is closer to ending, which relieves the pressure on the pitcher to avoid attempting a pickoff. The lead distance generally, although not strictly, increases as the number of outs increases. This makes intuitive sense because the runner is less likely to score without more aggressive action. In other words, the riskier lead distance is more palatable because the runner has a lower baseline probability of scoring from first base with two outs. The non-monotonicity of the optimal lead distance policy may be a consequence of estimating transition probabilities using empirical data.

@@

      \begin{table}[b]
        \centering
        \inputtable{lead_by_count_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by count and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with no outs and third base empty.
        }
        \label{tab:lead-by-count-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \inputtable{lead_by_outs_one_player.tex}
        \caption{
          \it The optimal lead distance (in feet) by outs and prior disengagements under the single-player MDP, assuming an average runner on first base facing an average battery (pitcher and catcher) with an 0-0 count and third base empty.
        }
        \label{tab:lead-by-outs-one-player}
      \end{table}

@@

      \begin{table}
        \centering
        \inputtable{lead_by_players.tex}
        \caption{
          \it Optimal lead distance (in feet) by battery/runner skill and prior disengagements, for an 0-0 count with no outs and third base empty. For example, the hypothetical 90$^{th}$-percentile battery represents the 90$^{th}$ percentile at attempting pickoffs, successfully executing pickoffs, suppressing stolen base attempts, and catching would-be base stealers.
        }
        \label{tab:lead-by-players}
      \end{table}

@@

    The next result states and proves the policy iteration algorithm for our two-player zero-sum sequential game.
       \begin{proposition}
           \label{lem:policy_iteration}
           Consider the sequential game defined in Section \ref{sec:sg-model} and suppose Assumptions \ref{assum:game_halts} and \ref{assum:transition_pby_continuous} hold. Starting from any $\pi_\text{R}^0 \in \Pi_\text{R}$, define the following updates for all $k \geq 0$.
           \begin{align*}
              % \label{eqn:update-value-two-player}
              V^{\pi_{\text{R}}^k}(s) &\defeq \min_{\pi_\text{P} \in \Pi_\text{P}} V^{\pi_\text{R}^k,\pi_\text{P}}(s) && \forall s \in \mathcal{S}, \mbox{ and}\\
              \label{eqn:update-policy-two-player}
              \pi_\text{R}^{k+1}(s) &\in \argmax_{a_\text{R} \in A_\text{R}(s)} \left\{\min_{a_\text{p} \in A_\text{P}(s,a_\text{R})}\sum_{s' \in \mathcal{S}}[r(s'|s) + V^{\pi_{\text{R}}^k}(s')] p(s'|s,a_\text{R},a_\text{P})\right\} && \forall s \in \mathcal{S}.
            \end{align*}
            Then $V^{\pi_\text{R}^k} \uparrow V^*$. 
       \end{proposition}
       \begin{proof}
           For any $\pi_\text{R} \in \Pi_\text{R}$, define the operator $T_{\pi_\text{R}}$ on $\mathbb{R}^{\abs{\cal S}}$ as 
           \begingroup
           \allowdisplaybreaks
           \begin{equation}
               \label{eqn:policy_iteration_temp1}
               (T_{\pi_\text{R}} u )(s) \defeq \min_{a_\text{P} \in A_\text{P}(s,\pi_\text{R}(s))}\sum_{s' \in S}\left[r(s'|s) + u(s')\right]p(s'|s,\pi_\text{R}(s),a_\text{P}) \ \forall s \in {\cal S}.
           \end{equation}
           \endgroup
           Since $\pi_\text{R}$ is stationary, the best response of the pitcher corresponding to $\pi_\text{R}$ reduces to solving a MDP. Denote the value of this MDP as $V^{\pi_\text{R}}$, i.e., $V^{\pi_\text{R}}(s) \defeq \min_{\pi_\text{P} \in \Pi_\text{P}} V^{\pi_\text{R},\pi_\text{P}}(s) \ \forall s \in {\cal S}$. Using similar arguments as in the proof of Theorem  \ref{lem:vi_convergence}, it follows that $\lim_{k \to \infty}T_{\pi_\text{R}}^k u = V^{\pi_\text{R}}$, where $T_{\pi_\text{R}}^k$ is the composition of $T_{\pi_\text{R}}$ taken with itself $k$ times.  Also, using arguments similar to the proof of Proposition  \ref{lem:bellman_equations}, it follows that $V^{\pi_\text{R}}$ is the unique fixed point of $T_{\pi_\text{R}}$.
